Sketch Recognition using Particle Swarm Optimizations

### 2009-05-26 paper2/Paper2.tex
change after moving to the new computer and before chaing it to doctor magdah request.

\documentclass{article}
\usepackage{spconf}
\usepackage{amsmath}
\usepackage[dvips,pdftex]{graphicx}
\usepackage{subfigure}
\usepackage{named}
\usepackage[dvips,pdftex]{hyperref}
\title{Sketch Recognition using Particle Swarm Optimizations}
\twoauthors
  {Maha El Meseery, Mahmoud Fakhr El Din\sthanks{Associate Professor}, Samia Mashali\sthanks{Professor}}
	{Signals Processing Group \\ Computers and Systems Department \\
  Electronic Research Institute\\Cairo, Egypt\\
	melmeseery@eri.sci.eg, samia@eri.sci.eg}
  {Magda Fayek\sthanks{Associate Professor}, Nevin Darwish\sthanks{Professor}}
	{ Computer Engineering Department\\
	Faculty of Engineering \\Cairo University\\Cairo, Egypt\\
	magdafayek@gmail.com,ndarwish@ieee.org}
\begin{document}
\maketitle
\begin{abstract}
Sketch recognition is defined as the process of identifying symbols that users draw using single or multiple strokes. Usually, users draw strokes using pens. The sketch recognition system immediately interprets their strokes as objects that can be easily manipulated. This paper uses Particle Swarm Optimization (PSO) to divide the strokes the user draws into meaningful geometric primitives. These geometric primitives are grouped to formulate symbols which are further identified. The results show that using PSO improves segmentation results which guide the symbol recognition phase. As for recognition we use Support Vector Machines (SVM) classifier which further improves the final recognition accuracy.  
\end{abstract}
\section{Introduction}
The field of symbol recognition has gained interest in the last few years. Scientists generally and engineers specifically express thoughts and designs using sketches. Engineers use sketches to exchange designs as a natural method of communication rather than writing or speaking. Design engineers need a powerful computer based system with symbol recognition capabilities to help them design, manipulate and store sketches more effectively than using only papers. Increasing the interaction between computers and users in sketch and CAD systems has been the main reason for the emerging of a few advanced sketch recognition systems.

Sketch recognition is divided into three main steps a) preprocessing, b) segmentation and c) symbol recognition. The preprocessing phase captures user input stroke points and collects basic information about the stroke. This is followed by proceeding to remove noise and computing basic statistical and geometrical information. In the segmentation phase, the strokes are divided into a set of simple geometrical primitives or segments. In the third phase, sketch recognition, strokes and segments are clustered to formulate symbols that can be recognized by a classifier.

This paper introduces a new sketch recognition system, which uses particle swarm optimization (PSO) to optimally segment user's strokes. Users can draw symbols using any number of strokes in any order they like. The system segments each stroke using two different PSO algorithms. The first segmentation algorithm is based on dividing the digital curve into polygons\cite{PolygonApproximationPSO}. Our algorithm enhances its performance by adding curvature and other local information while segmenting strokes to achieve better results. To handle complex strokes, our system uses a second segmentation PSO algorithm to segment strokes into a set of lines and circular arcs. After segmentation, a feature vector composed of different statistical, geometrical and composite features is deduced. The final classification step uses a SVM classifier to classify symbols into one of the previously trained classes based on the features vector computed. % uses the computed feature vector to classify the segments into one of the previously trained classes.

 The remaining of the paper is as follows; the literature review is in Section \ref{sec:review}. Section \ref{sec:ParticleSwarmAlgorithm} introduces Particle Swarm Optimization. Section \ref{Sysdisc} describes the system in details.  Sections \ref{Prepross}, \ref{seg} and \ref{sec:Recognition} provide details on preprocessing, segmentation and recognition process respectively. Section \ref{sec:Experiments} describes the experiments that were preformed. We conclude in Section \ref{ConclusionandFutureWork}.

\section{Literature Review}
\label{sec:review}
A wide variety of techniques were suggested for sketch recognition.  A hybrid algorithm was introduced in \cite{earlyprocess} where different sets of segments are generated based on both curvature and speed dominant points, followed by choosing a segmentation with the least error from a generated hybrid set. However, this system is limited to recognizing only specific simple geometric shapes with a set of low level recognizers. Each low level recognizer is designed to recognize only one geometric shape using spatial and geometric information extracted from input stroke.  Yu \cite{meanshift10} introduced a \textit{feature area} for each primitive and then computed the segmentation error for different types of primitives based on the computed \textit{feature area}. His system achieved good accuracy in simple shapes (square, ellipse,...etc) but did not perform well in complex shapes. Paulson and Hammond \cite{Paleosketch08} introduced a set of low level recognizers that were reported to achieve 98\%   accuracy. However, their system similar to all low level recognizers in the way that it identifies a small set of simple shapes.% A genetic algorithm was used by \cite{CruveDivisionSwarm} to optimally divide digital curves into lines and curves. Chen et al.\cite{CruveDivisionSwarm} uses digital curves scanned from paper as input to the system and did not take advantage of the curvature or local geometric properties of the digital curve.

A genetic algorithm was used by \cite{CruveDivisionSwarm} to optimally divide digital curves into lines and curves. Chen et al. \cite{CruveDivisionSwarm} uses digital curves scanned from paper as input to the system. Also, they did not take advantage of the curvature or local geometric properties of the digital curve. Yin \cite{PolygonApproximationPSO} used PSO to convert digital curves into polygons, our system adopts  Yin \cite{PolygonApproximationPSO} method and tries to improve it by adding curvature and other local information while segmenting strokes to achieve better segmentations. The next section presents the general particle swarm algorithm which is used in both our segmentation algorithms. % are presented in section \ref{subsubsec:Discreteparticleswarmalgorithm}

@@

\subsection{Preprocessing}
\label{Prepross}% Agar \cite{polygonfeedback31} mentions that the pointing device (for example the mouse) sampling rate is the reason for this phenomena.
 Speed and time difference information were widely used in sketch recognition systems \cite{earlyprocess}. Figure \ref{fig:speed2Distance} shows the time difference and the speed graphs for the stroke drawn in Figure \ref{fig:orignalStroke}. It was noted that time difference provides more distinct maxima than the speed information. Agar et al. \cite{polygonfeedback31} mention that the pointing device (for example the mouse) sampling rate is the reason for this phenomena. The points are sampled at regular time intervals as long as the pointing device is moving. Therefore, there are no new detected points when the pointing device is stationary. This leads to a nearly constant time difference between samples while the pen is moving and large difference while the pen is stationary (Figure \ref{fig:timediff}). Contrary to speed information, where the users draw with variant speeds, the speed information has a lot of noise in the data.  Similarly, direction information is used as it provides better distinctive maxima for the corners than the curvature information \cite{meanshift10}.% Figure \ref{fig:curvatures} shows the direction and curvature graphs for the stroke drawn in Figure \ref{fig:orignalStroke}
 \begin{figure}
	\centering
			\subfigure[ Speed Graph]{	\includegraphics[scale=0.4]{images/vel3.jpg}}
			\hfill
			\subfigure[ Time Difference Graph] {\label{fig:timediff}\includegraphics[scale=0.4]{images/td3.jpg}}
	\caption{Speed and Time Difference Graphs}
	\label{fig:speed2Distance}
\end{figure}
\begin{figure}[]
	\centering
		\includegraphics[scale=0.5]{images/stroke3.jpg}
	\caption{Input Stroke} Example of an input stroke to the system. 
	\label{fig:orignalStroke}
\end{figure}
Dominant points are characterized by lower speed values, higher curvature and direction values. However, the dominant points from the curvature information may not be the same points as the dominant points extracted from the speed information. Hence, we compute time difference, direction, speed and curvature of each point along a stroke to ensure that no dominant point is overlooked in this phase. The speed is calculated as $v=\Delta s/\Delta t$ where $t$ is the time difference between two points and $s$ is the length between them. The direction $d$ is calculated as the angle between two vectors and curvature is considered as the change in direction with respect to length i.e. $c= \Delta d/\Delta s$.

After the system computes the speed, time difference and curvature information it proceeds to detect the points with low velocity and high curvature. However, using simple differentiation to detect local extreme points resulted in false points due to the non smooth curves. Hence, the system adopted a process presented by \cite{earlyprocess}, where the mean of each of the four calcuated information  is calculated. This mean is taken as the threshold \textit{th} which is used to separate the curve into regions ($R_i$) (Figure\ref{fig:timediff}); each region $R_i$ is defined as a range of points, where the curve values are either above or below the threshold \textit{th}. Those regions are further processed to find the maximum point $Max(R_i)$ of each region $R_i$. The stroke points $p_i(x,y)$ that correspond to those maximum values are labeled as \textit{possible dominant points} $P_{pd}$. The system repeats this process for curvature, time difference and speed information. All the points labeled as possible dominant points $P_{pd}$ are saved into a single array. Figure \ref{fig:LabelsPPD} shows the particles labeled as Possible dominate point $P_{pd}$ as the preprocessing. It is clear that some of the $P_{pd}$ points are redundant as specified in the preprocessing stage. % (as shown are redundant)
\begin{figure}
	\centering
		\subfigure[Possible dominate point]{ 	\label{fig:LabelsPPD}	\includegraphics[scale=0.7]{images/ppd.jpg}}
		\hfill
			\subfigure[DPSO algorithm encoding]{	\label{fig:LabelsPSO} \includegraphics[scale=0.7]{images/pso1.jpg}}			
	\caption{Segmentation}% a) Possible dominate point b) Particle encoding  
	\label{fig:pso1}
\end{figure}

\subsection{Segmentation}
\label{seg}
In the segmentation stage the input stroke is divided into a set of primitives. As shown in Figure \ref{fig:segblock} first an attempt is made to fit the stroke points into a curve or an ellipse using a minimum square error fitting algorithm \cite{chernov-2003}. The ellipse fitting step helps to prevent the system from over segmenting the stroke into multiple lines or curves if the input stroke is an ellipse. If the stroke proved to be an ellipse arc then the segmentation process ends and the system proceeds to the next step. Otherwise, the stroke is passed to two furthur segmentation algorithms that divide the stroke to either lines or lines and curves. The next section describes in details the ellipse detection algorithm and the two PSO algorithms used to divide the stroke. % The input to these algorithms are the stroke points along with the possible dominant points $P_{pd}$ computed during preprocessing. The output is a set of dominant points which are connected by either lines or curves.

%///AllBlockDiagram.jpg  you can change the block to the main block. 
 \begin{figure}
	\centering
		\includegraphics[scale=0.5]{images/blockSmall.jpg}
	\caption{Segmentation Block}
	\label{fig:segblock}
\end{figure}
\subsection{Ellipse fitting} 
%The process starts by computing the center of the stroke bounding box. The bounding box center point is used as the first estimation of the ellipse center. The axes of the ellipse are estimated as $width/2$ and $height/2$ of the stroke bounding box. The least square fitting algorithm is used to minimize the fitting error of the ellipse equation \cite{chernov-2003}.  
t
This process tries to fit the stroke points into an ellipse arc; it starts with computing the center of the stroke bounding box. The bounding box center point is used as the first estimation of the center of the ellipse. The axes of the ellipse are estimated as the $width/2$ and $height/2$ of the stroke bounding box where $width$ is width and $height$ is height of the bonding box. The least square fitting algorithm \cite{chernov-2003} is used to minimize the fitting error of the ellipse Equation (\ref{eq:circleFit})

@@

\subsection{Non ellipse fitting algorithm}
\label{subsubsec:Discreteparticleswarmalgorithm}
Two DPSO algorithms are used to generate different segmentations for each stroke the user draws. For each segmentation generated an error is evaluated. The segmentation with the minimum error value is chosen as the best stroke segmentation (Figure \ref{fig:segblock}). The problem definition is the same in both algorithms but they differ in the method they use to compute fitness and error functions. 
\begin{description}
	\item[ Problem definition:] The input stroke $S$  with $N$ points can be represented by set $S = \left\{ {x_1 ,x_2  \ldots x_N }\right\}$ where $x_i$ is the location of the point $i$. The swarm algorithms consist of $M$ agents which are represented by the set $A = \left\{ {P_i \left| {i = 1,2 \cdots M} \right.} \right\}$ where $P_i$ is a single solution particle from the solution space. Each particle decodes the problem using a binary array with the same length $N$ as the input stroke.

Therefore, the system represents each particle $P_i$ by $P_i = \left\{ {p_{ij} \left| {j = 1,2 \cdots N} \right.} \right\}$ where $p_{ij}$ has only two values a)1 ($p_{ij}=1$); means that this point ($j$) is a dominate point, or b) 0($p_{ij}=0$) means this point ($j$) is not a dominate point. Figure\ref{fig:LabelsPSO} shows a particle encoding in the DPSO system.

@@

\section{Experimental}
\label{sec:Experiments}
The dataset collected by Hse and Newton\cite{HeloiseBeautification} is used to test the presented system. The data are drawn by 16 users each of them was asked to draw 13 shapes from 30 to 50 times. Figure \ref{fig:symbolSet} shows a set of the shapes used in the data set. The dataset was divided into training set and test set. Four different splits for the training and test data are generated from the dataset. The results displayed are the average recognition accuracy of the four splits. The accuracy is computed as the number of correctly recognized samples divided by the total number of samples in the test. %In the following section, the recognition accuracy means the average recognition accuracy in the four data splits.  
\begin{figure}
\centering 
\label{fig:symbolSet}
\fbox{ \parbox{5cm}{% 
		\includegraphics[scale=0.5]{images/symbolSet.PNG}	}}
		\caption{The Symbol Set} Symbols used in the dataset.
\end{figure}
\subsection{Algorithms comparisons}
\label{sec:AlgExp}
A set of test were performed on the system to enhance both the swarm algorithms outputs. The effect of number of agents, the error thresholds and various other paramters was investigated. The result of these tests are shown in Figure \ref{fig:swarmtesting} and   \ref{fig:swarmtesting2}. Figure \ref{fig:swarmtesting} shows that as the number of agents increases the error value decrease (error of the segmentation algorithm). The final number of agents and maximum swarm iteration used in the system was based trade off between error achieved versus the computation time. Other algorithm parameters like $c_1$,$c_2$ that were mentioned in Section\ref{sec:ParticleSwarmAlgorithm} are tested using similar tests (Figure \ref{fig:swarmtesting} and \ref{fig:swarmtesting2}).

 \begin{figure}
	\centering		
	 \includegraphics[scale=0.3]{images/swarmTesting.jpg}
	 	\caption{Swarm testing} Segmentation Error Vs. Number of agents
	 	\label{fig:swarmtesting}
	%\caption{Experiments results :} a)    b) c) 
\end{figure}

\begin{figure*}
	\centering		
	 \includegraphics[width=18cm]{images/swarmTesting2.jpg}
	 	\caption{Swarm testing} Segmentation Error Vs. PSO parameters.
	 	\label{fig:swarmtesting2}
	%\caption{Experiments results :} a)    b) c)

\end{figure*} 
We performed several experiments to evaluate the presented recognition system. Firstly we tested recognition accuracy of shapes in the data set with both algorithms (\textsl{AlgS1} and \textsl{AlgS2}). We also implemented the segmentation algorithm described in \cite{earlyprocess} (\textsl{Alg3}) to use as reference to compare it with our swarm algorithms. Figure \ref{fig:test1} shows the accuracy achieved by each algorithm. The two swarm algorithms were tested with and without the ellipse fitting module. The ellipse detection module appears to be superior. The result shows that combining both  (\textsl{AlgS1} and \textsl{AlgS2}) out perform any single algorithm. %show that combining algoirthmsachieve higher accuracy than any single algoirthm. 
%improves the results.% with both the \textit{DPSO} algorithms.% The results show that both \textit{DPSO} algorithms achieve better result than other algorithms.

 \begin{figure}
	\centering		
	 \includegraphics[scale=0.4]{images/testAlg.jpg}
	 	\caption{Algorithm comparison} The recognition rate of different algorithms. 
	 	\label{fig:test1}
	%\caption{Experiments results :} a)    b) c) 
\end{figure}

>>>>

\documentclass{article}
\usepackage{spconf}
\usepackage{amsmath}
\usepackage[dvips,pdftex]{graphicx}
\usepackage{subfigure}
\usepackage{named}
\usepackage[dvips,pdftex]{hyperref}
\title{Sketch Recognition using Particle Swarm Optimizations}
\twoauthors
  {Maha El Meseery\sthanks{Corresponding Author}, Mahmoud Fakhr El Din\sthanks{Associate Professor}, Samia Mashali\sthanks{Professor}}
	{Signals Processing Group \\ Computers and Systems Department \\
  Electronic Research Institute\\Cairo, Egypt\\
	melmeseery@eri.sci.eg, mafakhr@mcit.gov.eg,\\ samia@eri.sci.eg}
  {Magda Fayek\sthanks{Associate Professor}, Nevin Darwish\sthanks{Professor}}
	{Computer Engineering Department\\
	Faculty of Engineering \\Cairo University\\Cairo, Egypt\\
	magdafayek@gmail.com,ndarwish@ieee.org}
\begin{document}
\maketitle
\begin{abstract}
%This study 
Sketches and drawings are widely used as a simple method of expressing thought and designs. The goal of this paper is developing a sketch recognition system that facilitates using sketches in computer systems. The system uses Particle Swarm Optimization (PSO) algorithm to optimally segment the strokes the user draws into meaningful geometric primitives.  These geometric primitives are grouped to formulate symbols which are further identified using a Support Vector Machines (SVM) classifier. Two different segmentation algorithms were evaluated and results show that both algorithms improve the final recognition results. Experiments were conducted on three different dataset, simple presentation (Hs-DB), Electrical Design (EL-DB) and Logic Design (LD-DB) symbols. Results show that the effectiveness of the proposed system on various dataset especially on datasets with low shape diversity as in Logic Design dataset (LD-DS).  % sketch are divided  Users draw symbols and sketchs using a set of pen strokes. 
%Sketch recognition is defined as the process of identifying symbols that users draw using single or multiple strokes. Usually, users draw strokes using pens. The sketch recognition system immediately interprets their strokes as objects that can be easily manipulated. This paper uses Particle Swarm Optimization (PSO) to divide the strokes the user draws into meaningful geometric primitives. These geometric primitives are grouped to formulate symbols which are further identified. The results show that using PSO improves segmentation results which guide the symbol recognition phase. As for recognition we use Support Vector Machines (SVM) classifier which further improves the final recognition accuracy.  
\end{abstract}

\section{Introduction}

Computer research nowadays aims at facilitating computer human interaction in every way. Pen-based interfaces give the user a pencil-paper like feeling that enhances interaction more than the current used keyboard-mouse computer interfaces. However, Using sketches as input to a computer system requires the system to understand the drawn sketches. In a truly freely sketch system this is not a trivial problem due to the sloppy and messy way users draw sketches. Sketches here stand for any hand drawing the user draws using digital stylus or similar device. Normally, sketches are composed of a set of strokes. The time between pen-up and pen-down events identify each stroke. The strokes are mainly the path of points, which the user path through between the pen-down and pen-up events while using the stylus or similar device.

Sketch recognition can be based on: strokes, global shape properties, and appearance \cite{Oltmans07}. The role of which each stroke plays in the structure of the sketched symbol is used to determine shapes in stroke based methods. Most stroke based systems either represent symbols as gestures or as a  hierarchical shape. Gestures are pen strokes drawn by the user that can immediately recognized by the system. A Hierarchical shape descriptions was widely used in sketch systems \cite{HierarchicalParsing7,Ladder30,AlvaradoFreedom42,SketchRead2007}. Symbols are represented using hierarchical levels of simpler geometric shapes or primitives and spatial relations between them. Systems that employ such representation must first pre-process strokes to break them into geometric primitives. Recognition can then be treated as sub graph matching problem.

 In global shape properties approach, systems investigate general shape properties and its underlying strokes. Using shape properties (i.e. ratio of area of convex hull to area of shape) relaxes the assumption and role of each stroke in the symbol but it does not truly represent the appearance of the symbol\cite{DiagramOfflineConvexHull}. Unlike appearance based methods where individual strokes are disregarded and they focus on the overall appearance of shapes. In general, the  appearance based methods use image processing techniques to recognize sketches that the users draw.

 Methods based on gesture and global shape properties produce systems that are dependent on assumptions about how strokes can be grouped. These systems assume that users will finish drawing a shape before moving to the next one \cite{Cali63,geometrydomain49}. This let them to temporally group strokes and then recognizes them based on their strokes or global shape properties. \cite{AlvaradoDigital} argued that this assumption does not hold in natural sketches.

 Recognition based on strokes using hierarchical shape description is applied in \cite{SketchRead2007}. The principle problem in these systems is the computational complexity of the matching between shape description and geometric primitives produced in the segmentation process. Natural sketch are difficult to segment and group to create reliable fragment of geometric primitives. This is partially due to the increase of noise and other phenomena such as overlapped, over traced strokes and sharing of strokes between different symbols. Since these systems are dependable on the segmentation results, they either perform recognition using undependable segmentation results or consider different ways to segment the input sketch.

In this paper we choose to implement a hybrid method between shape property and stroke based approaches. Our method tries to avoid many of the problems faced by approaches based on individual strokes or global shape properties while trying to gain the additional information that the stroke based system use. Firstly, the system uses particle swarm optimization (PSO) to optimally segment user's strokes. Users can draw symbols using any number of strokes in any order they like. The system segments each stroke using two different PSO algorithms. The first segmentation algorithm is based on dividing the digital curve into polygons\cite{PolygonApproximationPSO}. Our algorithm enhances its performance by adding curvature and other local information while segmenting strokes to achieve better results. To handle complex strokes, our system uses a second segmentation PSO algorithm to segment strokes into a set of lines and circular arcs. After segmentation, a feature vector composed of different statistical, geometrical and composite features is deduced. The final classification step uses a SVM classifier to classify symbols into one of the previously trained classes based on the features vector computed.

 %The preprocessing phase captures user input stroke points and collects basic information about the stroke. In the segmentation phase, the strokes are divided into a set of simple geometrical primitives or segments. In the third phase, sketch recognition, strokes and segments are clustered to formulate symbols that can be recognized by a classifier.

%This paper introduces a new sketch recognition system, which uses  % uses the computed feature vector to classify the segments into one of the previously trained classes.

 The remaining of the paper is as follows; the literature review on stroke segmentation is in Section \ref{sec:review}. Section \ref{sec:ParticleSwarmAlgorithm} introduces Particle Swarm Optimization. Section \ref{Sysdisc} describes the system in details.  Sections \ref{Prepross}, \ref{seg} and \ref{sec:Recognition} provide details on preprocessing, segmentation and recognition process respectively. Section \ref{sec:Experiments} describes the experiments that were preformed. We conclude in Section \ref{ConclusionandFutureWork}.

\section{Literature Review}
\label{sec:review}

Yu \cite{meanshift10} introduced a \textit{feature area} for each primitive and then computed the segmentation error for different types of primitives based on the computed \textit{feature area}. His system achieved good accuracy in simple shapes (square, ellipse,...etc) but did not perform well in complex shapes. A hybrid algorithm was introduced in \cite{earlyprocess} where different sets of segments are generated based on both curvature and speed dominant points, followed by choosing a segmentation with the least error from a generated hybrid set. However, this system is limited to recognizing only specific simple geometric shapes with a set of low level recognizers. Each low level recognizer is designed to recognize only one geometric shape using spatial and geometric information extracted from input stroke.

A genetic algorithm was used by \cite{CruveDivisionSwarm} to optimally divide digital curves into lines and curves. Chen et al. \cite{CruveDivisionSwarm} uses digital curves scanned from paper as input to the system. Also, they did not take advantage of the curvature or local geometric properties of the digital curve. Yin \cite{PolygonApproximationPSO} used PSO to convert digital curves into polygons, our system adopts  Yin \cite{PolygonApproximationPSO} method and tries to improve it by adding curvature and other local information while segmenting strokes to achieve better segmentations. The next section presents the general particle swarm algorithm which is used in both our segmentation algorithms.

 % are presented in section \ref{subsubsec:Discreteparticleswarmalgorithm}

@@

\subsection{Preprocessing}
\label{Prepross}% Agar \cite{polygonfeedback31} mentions that the pointing device (for example the mouse) sampling rate is the reason for this phenomena.
 Speed and time difference information were widely used in sketch recognition systems \cite{earlyprocess}. Figure \ref{fig:speed2Distance} shows the time difference and the speed graphs for the stroke drawn in Figure \ref{fig:orignalStroke}. It was noted that time difference provides more distinct maxima than the speed information. Agar et al. \cite{polygonfeedback31} mention that the pointing device (for example the mouse) sampling rate is the reason for this phenomena. The points are sampled at regular time intervals as long as the pointing device is moving. Therefore, there are no new detected points when the pointing device is stationary. This leads to a nearly constant time difference between samples while the pen is moving and large difference while the pen is stationary (Figure \ref{fig:timediff}). Contrary to speed information, where the users draw with variant speeds, the speed information has a lot of noise in the data.  Similarly, direction information is used as it provides better distinctive maxima for the corners than the curvature information \cite{meanshift10}.% Figure \ref{fig:curvatures} shows the direction and curvature graphs for the stroke drawn in Figure \ref{fig:orignalStroke}
 \begin{figure}[h]
	\centering
		\includegraphics[scale=0.5]{images/stroke3.jpg}
	\caption{Input Stroke} Example of an input stroke to the system. 
	\label{fig:orignalStroke}
\end{figure}
 \begin{figure}
	\centering
			\subfigure[ Speed Graph]{	\includegraphics[scale=0.4]{images/vel3.jpg}}
			\hfill
			\subfigure[ Time Difference Graph] {\label{fig:timediff}\includegraphics[scale=0.4]{images/td3.jpg}}
	\caption{Speed and Time Difference Graphs}
	\label{fig:speed2Distance}
\end{figure}
Dominant points are characterized by lower speed values, higher curvature and direction values. However, the dominant points from the curvature information may not be the same points as the dominant points extracted from the speed information. Hence, we compute time difference, direction, speed and curvature of each point along a stroke to ensure that no dominant point is overlooked in this phase. The speed is calculated as $v=\Delta s/\Delta t$ where $t$ is the time difference between two points and $s$ is the length between them. The direction $d$ is calculated as the angle between two vectors and curvature is considered as the change in direction with respect to length i.e. $c= \Delta d/\Delta s$.

After the system computes the speed, time difference and curvature information it proceeds to detect the points with low velocity and high curvature. However, using simple differentiation to detect local extreme points resulted in false points due to the non smooth curves. Hence, the system adopted a process presented by \cite{earlyprocess}, where the mean of each of the four calculated information  is calculated. This mean is taken as the threshold \textit{th} which is used to separate the curve into regions ($R_i$) (Figure\ref{fig:timediff}); each region $R_i$ is defined as a range of points, where the curve values are either above or below the threshold \textit{th}. Those regions are further processed to find the maximum point $Max(R_i)$ of each region $R_i$. The stroke points $p_i(x,y)$ that correspond to those maximum values are labeled as \textit{possible dominant points} $P_{pd}$. The system repeats this process for curvature, time difference and speed information. All the points labeled as possible dominant points $P_{pd}$ are saved into a single array. Figure \ref{fig:ppd999} shows the particles labeled as Possible dominate point $P_{pd}$ as the preprocessing. It is clear that some of the $P_{pd}$ points are redundant as specified in the preprocessing stage. % (as shown are redundant)
\begin{figure}
	\centering
\includegraphics[scale=0.7]{images/ppd.jpg}
	\caption{Possible Dominante Points}
	%\label{fig:ppd}
	\label{fig:ppd999}	
\end{figure}

\subsection{Segmentation}
\label{seg}
In the segmentation stage the input stroke is divided into a set of primitives. As shown in Figure \ref{fig:segblock} first an attempt is made to fit the stroke points into a curve or an ellipse using a minimum square error fitting algorithm \cite{ellipsefit}. The ellipse fitting step helps to prevent the system from over segmenting the stroke into multiple lines or curves if the input stroke is an ellipse. If the stroke proved to be an ellipse arc then the segmentation process ends and the system proceeds to the next step. Otherwise, the stroke is passed to two further segmentation algorithms that divide the stroke to either lines or lines and curves. After two different segmentations are generated, the system chooses the segmentation with the minimum error. The following sections describe in details the ellipse detection algorithm and the two segmentation algorithms used to divide the stroke. % The input to these algorithms are the stroke points along with the possible dominant points $P_{pd}$ computed during preprocessing. The output is a set of dominant points which are connected by either lines or curves.

%///AllBlockDiagram.jpg  you can change the block to the main block. 
 \begin{figure}
	\centering
		\includegraphics[scale=0.48]{images/blockSmall.jpg}
	\caption{Segmentation Block}
	\label{fig:segblock}
\end{figure}
\subsection{Ellipse fitting}
%The process starts by computing the center of the stroke bounding box. The bounding box center point is used as the first estimation of the ellipse center. The axes of the ellipse are estimated as $width/2$ and $height/2$ of the stroke bounding box. The least square fitting algorithm is used to minimize the fitting error of the ellipse equation \cite{chernov-2003}.  
This process tries to fit the stroke points into an ellipse arc; it starts with computing the center of the stroke bounding box. The bounding box center point is used as the first estimation of the center of the ellipse. The axes of the ellipse are estimated as the $width/2$ and $height/2$ of the stroke bounding box where $width$ is width and $height$ is height of the bonding box. The least square fitting algorithm \cite{chernov-2003} is used to minimize the fitting error of the ellipse Equation (\ref{eq:circleFit})

@@

\subsection{Non ellipse fitting algorithm}
\label{subsubsec:Discreteparticleswarmalgorithm}
\begin{figure}
	\centering
		%\subfigure[Possible dominate point]{ 	\label{fig:LabelsPPD}	\includegraphics[scale=0.7]{images/ppd.jpg}}
	%	\hfill
	 \includegraphics[scale=0.7]{images/pso1.jpg}			
	\caption{DPSO algorithm encoding}% a) Possible dominate point b) Particle encoding  
	\label{fig:pso1}
\end{figure}
Two DPSO algorithms are used to generate different segmentations for each stroke the user draws. For each segmentation generated an error is evaluated. The segmentation with the minimum error value is chosen as the best stroke segmentation (Figure \ref{fig:segblock}). The problem definition is the same in both algorithms but they differ in the method they use to compute fitness and error functions. 
\begin{description}
	\item[ Problem definition:] The input stroke $S$  with $N$ points can be represented by set $S = \left\{ {x_1 ,x_2  \ldots x_N }\right\}$ where $x_i$ is the location of the point $i$. The swarm algorithms consist of $M$ agents which are represented by the set $A = \left\{ {P_i \left| {i = 1,2 \cdots M} \right.} \right\}$ where $P_i$ is a single solution particle from the solution space. Each particle decodes the problem using a binary array with the same length $N$ as the input stroke.

Therefore, the system represents each particle $P_i$ by $P_i = \left\{ {p_{ij} \left| {j = 1,2 \cdots N} \right.} \right\}$ where $p_{ij}$ has only two values a)1 ($p_{ij}=1$); means that this point ($j$) is a dominate point, or b) 0($p_{ij}=0$) means this point ($j$) is not a dominate point. Figure\ref{fig:pso1} shows a particle encoding in the DPSO system.

@@

\section{Experimental}
\label{sec:Experiments}
In our experiment we used three different datasets. First dataset was collected by Hse and Newton\cite{HeloiseBeautification} (Hs-DB). The data are drawn by 16 users each of them was asked to draw 13 shapes from 30 to 50 times. Figure \ref{fig:symbolSet} shows a set of the shapes used in the data set. We collected the second and third datasets. Figure \ref{fig:ELsymbolSet} and \ref{fig:LogicsymbolSet} shows the symbols gathered from 7 different users in electrical (EL-DB) and logic design (LD-DB) domain respectively.  Table \ref{tab:datasets} shows a  comparison between the three datasets used in our experiments. The datasets were divided into training set and test set. Five different splits  for the training and test data are generated from each dataset. The results displayed are the average recognition accuracy of the five splits. The accuracy is computed as the number of correctly recognized samples divided by the total number of samples in the test.

\begin{figure}
\centering 
\label{fig:symbolSet}
\fbox{ \parbox{5cm}{% 
		\includegraphics[scale=0.5]{images/symbolSet.PNG}	}}
		\caption{The Hs-DB Symbol Set}.
\end{figure}
\begin{figure}
\centering 
\label{fig:LogicsymbolSet}

		\includegraphics[scale=0.5]{images/logicSet.jpg}	 
		\caption{The Digital Design Symbols Dataset(LD-DB)} Symbols used in the dataset.
\end{figure}
\begin{figure}
\centering 
\label{fig:ELsymbolSet}
\fbox{ \parbox{4.5cm}{% 
		\includegraphics[scale=0.5]{images/EelectImage.jpg}	}}
		\caption{The Electrical Symbols Dataset(EL-DB)} .
\end{figure}

\begin{table}
\begin{center}
\scalebox{0.9}{
 \begin{tabular}{|p{2cm}|p{1.5cm}|p{1.5cm}|p{1.5cm}|}
 \hline
 & Hs-DB &  EL-DB & LD-DB \\ \hline
No of samples  &  7791 & 2764 & 1859 \\ \hline
No of categories  & 13 & 15 & 8 \\ \hline
Avg. Samples per Category& 600 & 184 & 232 \\\hline 
No of users  & 16 & 7 & 7 \\  \hline
Balanced & Yes&No & No \\ \hline
Splits  & 5 random splits  & 5 random splits  & 5 random splits  \\ \hline
Examples & Figure \ref{fig:symbolSet}  & Figure \ref{fig:ELsymbolSet}  & Figure \ref{fig:LogicsymbolSet}   \\ \hline
\end{tabular}
}
\caption[Datasets Comparisons]{Datasets Comparisons} 
\label{tab:datasets}

\end{center}
\end{table}

%The dataset collected by Hse and Newton\cite{HeloiseBeautification} is used to test the presented system. The data are drawn by 16 users each of them was asked to draw 13 shapes from 30 to 50 times. Figure \ref{fig:symbolSet} shows a set of the shapes used in the data set. The dataset was divided into training set and test set. Four different splits for the training and test data are generated from the dataset. The results displayed are the average recognition accuracy of the four splits. The accuracy is computed as the number of correctly recognized samples divided by the total number of samples in the test. %In the following section, the recognition accuracy means the average recognition accuracy in the four data splits.

\subsection{Algorithms comparisons}
\label{sec:AlgExp}
A set of tests were performed on the system to enhance non ellipse fitting algorithm. The effect of number of agents, the error thresholds and various other parameters were investigated. The result of these tests are shown in Figure \ref{fig:swarmtesting} and   \ref{fig:swarmtesting2}. Figure \ref{fig:swarmtesting} shows that as the number of agents increases the error value decrease (error of the segmentation algorithm). The final number of agents and maximum swarm iteration used in the system was based on a trade off between error achieved versus the computation time. Other algorithm parameters such as $c_1$,$c_2$ that were mentioned in Section\ref{sec:ParticleSwarmAlgorithm} are tested using similar tests (Figure \ref{fig:swarmtesting} and \ref{fig:swarmtesting2}).

 \begin{figure}
	\centering		
	 \includegraphics[scale=0.3]{images/swarmAgentTest.jpg}
	 	\caption{Swarm testing} Segmentation Error Vs. Number of agents
	 	\label{fig:swarmtesting}
	%\caption{Experiments results :} a)    b) c) 
\end{figure}

\begin{figure}
	\centering		
	 \includegraphics[scale=0.4]{images/swarmParamterC.jpg}
	 	\caption{Swarm testing} Segmentation Error Vs. PSO parameters.
	 	\label{fig:swarmtesting2}
	%\caption{Experiments results :} a)    b) c)

\end{figure}

We performed several experiments to evaluate the presented recognition system. Firstly we tested recognition accuracy of shapes in the data set with both algorithms (\textsl{AlgS1} and \textsl{AlgS2}). We also implemented the segmentation algorithm described in \cite{earlyprocess} (\textsl{Alg3}) to use as reference to compare it with our swarm algorithms. Figure \ref{fig:test1} and \ref{fig:testLD} shows the accuracy achieved by each algorithm. The two swarm algorithms were tested with and without the ellipse fitting module. The ellipse detection module appears to be superior. The result shows that combining both  (\textsl{AlgS1} and \textsl{AlgS2}) out perform any single algorithm. %show that combining algoirthmsachieve higher accuracy than any single algoirthm. 
%improves the results.% with both the \textit{DPSO} algorithms.% The results show that both \textit{DPSO} algorithms achieve better result than other algorithms.
